Hyper-linked Communications: WebRTC enabled asynchronous collaboration

### 2016-03-12 tese1/08_related_work_rtc.tex
[WR] 1st part corrections

\subsection{Early days of the Internet and its remaining flaws}\label{early}

The need to build a global communications network in an age when almost nobody had access to that technology and the number of future users was unpredictable, lead to some protocols not being suitable for the huge growth of users that followed. \ac{IPv4} limits the number of public addresses in such a way that today they are scarce \cite{ipv4}. One way to overcome this problem was the development of a mechanism that groups multiple address into a single one, the machine that is assigned that address is then responsible for redirecting messages to members of its group using their private addresses, each connection in the private network is identified publicly by the same \ac{IP} address with a different port.
%RP o porto não identifica uma máquina (membro da rede privada) mas uma ligação.
This technique is known as \ac{NAT}.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.4\textwidth]{figures/nat.png}
	\caption{Network Address Translation}
\end{figure}
%RP na figura devias indicar que uns tem ips publicos e outros privados. Alternativamente, colocavas metade da nat box dentro da núvem para indicar que esta pertence à Internet e os pc não.

Initially \ac{NAT} offered an alternative to address exhaustion and a minimal sensation of security, although its current wide usage, \ac{NAT}s are exposing their weaknesses to the application layer.
%RP initially? E hoje?
There are four types of \ac{NAT} implementations\cite{rfc3489}: \emph{Full Cone NAT}, \emph{Restricted Cone NAT}, \emph{Port Restricted Cone NAT}, \emph{Symmetric NAT}.
%RP apresentas 5. falas em asymmetric. É um erro?

\emph{Full Cone} \ac{NAT} maps each public \ac{IP} address and port to a private \ac{IP} address and port.
Any external host can communicate with private hosts through their mapped public address and port. This represents the least restrictive type of \ac{NAT} and as we will later, the unique type of \ac{NAT} that enables real time communications from point to point.

\emph{Restricted Cone} \ac{NAT} requires that a private client must first send a message to an external host before it can receive messages from the same host. With this type of \ac{NAT}, the private client can be contacted from any port of the same external host.

\emph{Port Restricted Cone} \ac{NAT} works in the same way as Restricted Cone \ac{NAT}, but it only allows communications from the same external host's IP address and port, ignoring all messages from other applications within the same external host.

\emph{Symmetric} NAT maps different ports for each connection, as we will see later, this type of \ac{NAT} represents a problem on real time communications.

\emph{Non-Symmetric} \ac{NAT}s became the common configurations on the Internet. As a direct result, problems started to appear: the amount of ports that \ac{IP} makes available is also small compared to our current needs; worse than that, \ac{NAT} also difficults end-to-end communication, forcing most applications that follow this model to be implemented ineffectively.

Applications behind a \ac{NAT} are prevented to receive incoming connections from the public network, which forces them to behave as a client of a client-server model.

%RP explicar que o nat funciona num cenário em que a máquina atrás do nat é apenas um consumidor de informação, trabalhando em modo cliente do modelo cliente-servidor. Em P2P isto falha.

Applications based on multimedia and peer-to-peer file sharing have been one of the most strained by \ac{NAT}. Those kind of applications require real time communication in order to achieve the best performance.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.30\textwidth]{figures/multinat.png}
	\caption{Multiple level \ac{NAT}s}
\end{figure}

%RP file sharing não precisa de real-time
\ac{STUN} and \ac{TURN} \cite{natvoip} servers are a possible solution to overcome \ac{NAT}, although, none of those can establish direct connections on multiple level \ac{NAT}s.
%HR XXX I was here
%RP explicar o que são multiple level pois não falaste disto antes? São ``nested''?

\ac{STUN} servers are quite simple. They receive requests from \ac{NAT}ed clients, with the source address of a request being the public address that \ac{NAT} mapped to the client. \ac{STUN} servers will then reply to the client, providing the mapped public address, so it knows its associated public \ac{IP} address and port. Symmetric \ac{NAT} changes \ac{IP} port for each different connection, for that reason, when the \ac{STUN} servers reply with the \ac{IP} address and port of their connection, it will be useless for clients to use in other connection connections. That is why Symmetric \ac{NAT} represents a problem for peer-to-peer communications.   
%RP os problemas que relatas têm a ver com P2P ou real-time?

\ac{TURN} uses public servers to relay traffic between private endpoints.
It may use a \ac{P2P} network relay to find the best peer, but after that, the behavior is much like client-server. Direct communication is only achieved by \ac{STUN} when \ac{NAT} is a type \emph{full cone}. \ac{ICE} is a technique that uses \ac{STUN} when direct communications are possible and \ac{TURN} while a direct communication isn't possible.
%RP ainda não explicaste o que é o ICE
%RP explicar que para usar TURN as aplicações têm de estar cientes disso.

Most of client-server applications aren't affected by \ac{NAT} when the servers are public, but they're inadequate for real time communication between two private endpoints. Clearly \ac{TURN} requires a more expensive relaying infrastructure and, in most cases, more network usage, leading to a worse quality of service. The requirements of real-time video communication makes this kind of model unsuitable.
%RP quando falas em infraestrutura mais dispendiosa para o video, está a falar da internet ou ainda estamos a falar de usar ICE?

When connection is established, either in a direct or indirect way (via \ac{TURN} servers), \ac{WebRTC} came to simplify how audio and video are transmitted through web browsers.

@@

Some operating systems such as \emph{Android}, \emph{iOS}, \emph{Linux}, \emph{OSX} and \emph{Windows} implements native \ac{WebRTC} libraries, extending the usage of \ac{WebRTC} to applications outside the web browser. This native support can help to implement applications that record video and audio streams for further playback.

@@

\ac{TLS} doesn't support independent record decryption, for that it requires a reliable transport channel, typically \ac{TCP}. The decryption of a record depends on the previous record, which for unreliable transport protocols like \ac{UDP} may represent a problem, either due to packet loss or different reception order.

>>>>

@@

Some operating systems such as \emph{Android}, \emph{iOS}, \emph{Linux}, \emph{OSX} and \emph{Windows} implement native \ac{WebRTC} libraries, extending the usage of \ac{WebRTC} to applications outside the web browser. This native support can help to implement applications that record video and audio streams for further playback.

@@

\ac{TLS} doesn't support independent packet decryption\cite{rfc6347}, for that it requires a reliable transport channel, typically \ac{TCP}. The decryption of a packet depends on the previous packet, which for unreliable transport protocols like \ac{UDP} may represent a problem, either due to packet loss or different reception order.

### 2016-03-12 tese1/10_related_work_hypermedia.tex
[WR] 1st part corrections

  SmillingWeb \cite{smillingweb} attempts to implement a cross platform multimedia player designed for \ac{SMIL} 3.0 presentations with \emph{JavaScript} and \emph{jQuery} which, unlike \cite{ambulant}, doesn't require a plug-in to be installed and shouldn't have incompatibility issues. 
  SmillingWeb already takes advantage of \ac{HTML}5 and \ac{CSS}3.
  It takes into account unsupported web browsers through the use of \emph{Modernizr}\footnote{\url{http://modernizr.com/}}, a simple \emph{JavaScript} library that may require plug-ins if new features aren't supported.  
  But SmillingWeb just implements a subset of \ac{SMIL} 3.0 and their scheduler engine loads the \ac{SMIL} file only once, which could raise problems when dealing with \ac{SMIL} changes due to real time communications.
  Another problem with SmillingWeb is pre-loading and playing elements at the correct interval of time, which is not always possible due to high latency networks leading to  pauses during playback.

@@

  For example, with \ac{API}s like WebGL\footnote{\url{http://khronos.org/webgl/}}, it is now possible to manipulate a three dimensional environment in the context of a hyper-call. Another example would be a collaborative spreadsheet using \ac{WebRTC}. With this, hyper-calls are not limited to only audio, image, text and video, but also interaction with complex graphical user interfaces that changes over time.
 %XXX
  %RP porque escolheste falar aogra de webgl se não falas antes?
%HR já troquei, o WebGL usa o canvas do HTML5, talvez seja este o local mais adequado

@@

  Back in 1995, \emph{flash}\footnote{\url{http://www.adobe.com/products/flash.html}} was developed for web-based animations. Introducing video support in 2002, \emph{flash} started to grow after that. Competitor's players, at that time were focused on playing video and audio, while \emph{flash} had vector graphics and focused on streaming \emph{on-demand} video across multiple platforms. \emph{VP6} was their choice of video codec, providing half the video size (in Byte) for the same quality and providing video quality adjusted to \emph{Internet} connection latency. 
  In 2010, \emph{Adobe Flash} was the most widely used applications for reproducing live broadcast and recorded video \cite{flashvideo}, it supports progressive video download using \ac{HTTP} and streaming using \ac{RTMP}.
  %RP ``was'' quando? Em 2010? Ontem?
  %HR exacto.

@@

  Like \emph{Flash}, \emph{Microsoft Silverlight}\footnote{\url{http://www.microsoft.com/silverlight/}} is a cross browser plug-in and platform that is used to develop rich \emph{Internet} applications. It supports vector graphics, animation and video. Compared to \emph{flash}, which uses \emph{ActionScript}, \emph{Silverlight} applications can use languages like C\#, \emph{VisualBasic} and \ac{XAML}. \emph{Silverlight} uses a technique called \emph{Smooth Streaming} from \emph{IIS Media Service} that consists on delivering video in real-time with adjusted quality in function of bandwidth variations and \ac{CPU} usage.

>>>>

  SmillingWeb \cite{smillingweb} attempts to implement a cross platform multimedia player designed for \ac{SMIL} 3.0 presentations with \emph{JavaScript} and \emph{jQuery} which, unlike \cite{ambulant}, doesn't require a plug-in to be installed and shouldn't have incompatibility issues. 
  SmillingWeb already takes advantage of \ac{HTML}5 and \ac{CSS}3.
  It takes into account unsupported web browsers through the use of \emph{Modernizr}\footnote{\url{http://modernizr.com/}(accessed June 2, 2015).}, a simple \emph{JavaScript} library that may require plug-ins if new features aren't supported.  
  But SmillingWeb just implements a subset of \ac{SMIL} 3.0 and their scheduler engine loads the \ac{SMIL} file only once, which could raise problems when dealing with \ac{SMIL} changes due to real time communications.
  Another problem with SmillingWeb is pre-loading and playing elements at the correct interval of time, which is not always possible due to high latency networks leading to  pauses during playback.

@@

  For example, with \ac{API}s like WebGL\footnote{\url{http://khronos.org/webgl/}(accessed June 2, 2015).}, it is now possible to manipulate a three dimensional environment in the context of a hyper-call. Another example would be a collaborative spreadsheet using \ac{WebRTC}. With this, hyper-calls are not limited to only audio, image, text and video, but also interaction with complex graphical user interfaces that changes over time.
 %XXX
  %RP porque escolheste falar aogra de webgl se não falas antes?
%HR já troquei, o WebGL usa o canvas do HTML5, talvez seja este o local mais adequado

@@

  Back in 1995, \emph{flash}\footnote{\url{http://www.adobe.com/products/flash.html}(accessed June 2, 2015).} was developed for web-based animations. Introducing video support in 2002, \emph{flash} started to grow after that. Competitor's players, at that time were focused on playing video and audio, while \emph{flash} had vector graphics and focused on streaming \emph{on-demand} video across multiple platforms. \emph{VP6} was their choice of video codec, providing half the video size (in Byte) for the same quality and providing video quality adjusted to \emph{Internet} connection latency. 
  In 2010, \emph{Adobe Flash} was the most widely used applications for reproducing live broadcast and recorded video \cite{flashvideo}, it supports progressive video download using \ac{HTTP} and streaming using \ac{RTMP}.
  %RP ``was'' quando? Em 2010? Ontem?
  %HR exacto.

@@

  Like \emph{Flash}, \emph{Microsoft Silverlight}\footnote{\url{http://www.microsoft.com/silverlight/}(accessed June 2, 2015).} is a cross browser plug-in and platform that is used to develop rich \emph{Internet} applications. It supports vector graphics, animation and video. Compared to \emph{flash}, which uses \emph{ActionScript}, \emph{Silverlight} applications can use languages like C\#, \emph{VisualBasic} and \ac{XAML}. \emph{Silverlight} uses a technique called \emph{Smooth Streaming} from \emph{IIS Media Service} that consists on delivering video in real-time with adjusted quality in function of bandwidth variations and \ac{CPU} usage.

### 2016-03-12 tese1/11_related_work_collaboration.tex
[WR] 1st part corrections

	\emph{ShareJS}\footnote{\url{http://sharejs.org/}(accessed June 2, 2015).} is an \ac{OT} \emph{JavaScript} library, developed by the ex \emph{Google Wave} engineer Joseph Gentle, for collaborative text and \ac{JSON} documents edition in real-time.

	\emph{TogetherJS}\footnote{\url{http://togetherjs.com/}(accessed June 2, 2015).} is a JavaScript library that uses \ac{WebRTC} for collaborative web applications. It uses \ac{JSON} messages for \ac{OT} concurrency control but it does not provide storage.

@@

>>>>

	\emph{ShareJS}\footnote{\url{http://sharejs.org/}(accessed June 2, 2015).} is an \ac{OT} \emph{JavaScript} library, developed by the ex \emph{Google Wave} engineer Joseph Gentle, for collaborative text and \ac{JSON} documents edition in real-time. {\color{blue} It uses \emph{ShareDB} \footnote{\url{https://github.com/share/sharedb}(accessed March 10, 2016)} for its backend and data model, which supports simple integration with any database. One of \emph{ShareDB}'s integration is over \emph{MongoDB}\footnote{\url{https://github.com/share/sharedb-mongo}(accessed March 10, 2016)}}.

	\emph{TogetherJS}\footnote{\url{http://togetherjs.com/}(accessed June 2, 2015).} is a JavaScript library that uses \ac{WebRTC} for collaborative web applications. It uses \ac{JSON} messages for \ac{OT} concurrency control but it does not provide storage. {\color{blue}\emph{TogetherJS} uses its own servers for the signaling phase and it supports microphone and mouse sharing between users.}

@@

	{\color{blue}\emph{OT.js}\footnote{\url{http://operational-transformation.github.io}(accessed March 10, 2016)} is a \emph{JavaScript} library that only implements operation transformations on client side over plain text. An implication of implementing just the client side is the extra effort that is necessary to implement content's persistent storage, besides this drawback this library is very flexible because it's not tied to a specific database.}

{\color{green} For ease of development we decided to use the same backend for \ac{OT} since our web server relies on \emph{Play Framework} with Java and \emph{ShareJS} relies on \emph{NodeJS}\footnote{\url{https://nodejs.org/en/(accessed March 10, 2016)}} }

{\color{green} \emph{TogetherJS} requires a simple server (also known as hub) that echoes messages between clients, all the synchronization work is performed by the clients. Altough the hub's reduced complexity, \emph{TogetherJS}'s server is implemented with \emph{NodeJS}, we would need read the hub's source code and understand what changes are needed to perform on our web server in order to make it compatible with \emph{TogetherJS}.}

### 2016-03-12 tese1/project.tex
[WR] 1st part corrections

\begin{abstract}
The Hyper-linked communications concept applies much of the hypermedia concepts, widely used on Web content, to communication and collaboration services. This paradigm allows to synchronize, structure and navigate communication content, providing a way to integrate social media content and collaborative tools into voice and video calls.
Voice and image together can express emotions and expose creativity like no other medium can. With hypermedia concepts, we can add more value to voice and video communications.

\ac{WebRTC} technology allows real-time communications between web browsers without the need to install and use additional applications or plug-ins. The nature of web browser applications already follow the hypermedia concept, which makes \ac{WebRTC} the ideal technology to apply the hyper-linked communications concepts.
The native support of \ac{WebRTC} in operating systems extends its usage to outside the web browser, allowing to explore functionalities that browsers are poor to support such as video recording and massive information storage.

Our goal in this project is to develop an application that leverages the Hyper-linked communications concept to enable collaborative real-time audio and video communication, enriched with other media types, that can be accessed live or accessed latter.
This application will target the web browser platform, resorting to \ac{WebRTC}.

In this document, we present the current State Of The Art in hyper-linked communications and related technologies, propose an architecture to implement an hyper-linked communication application based on \ac{WebRTC} and provide a plan for its implementation and evaluation.

\keywords{WebRTC, asynchronous, communications, collaboration}
\end{abstract}

\section{Introduction}
  \input{intro.tex}

\section{Related Work}\label{related}
 This section is structured as follows.
 Section \ref{early} describes the problems that real-time communications face on nowadays internet, namely the \ac{IPv4} address exhaustion and the client server model constraints. 
 Section \ref{rtc} describes the \ac{WebRTC} technology and the protocols needed to implement our project. 
 Section \ref{signaling} addresses the signaling component of chat applications, which is not defined on \ac{WebRTC} specifications. 
 Section \ref{hypermedia} presents the evolution of multimedia content until the hypermedia, its capabilities, synchronization mechanisms and interactivity. 
 Section \ref{collab} explores streaming protocols for non-interactive multimedia and how to introduce the interactive component, another important aspects are the ability to control the time flux of a stream and collaborative application development.
%RP este detalhe sobre a estrutura da sec. do estado da arte vai para o início dessa secção.

  \input{state01.tex}
  \input{state02.tex}
  \input{state03.tex}
  \input{state04.tex}

@@

%\subsection{Summary} % English
%Neste secção deve-se fazer o resumo do trabalho efectuado, retomando a ideia, as contribuições definidas e a forma como estas se materializaram

%\subsection{Conclusions} % English
%Nesta secção devem ser retiradas conclusões do trabalho realizado, em face dos resultados obtidos \cite{exemplo}.

>>>>

\input{01_abstract.tex}
\input{02_introduction.tex}
\input{03_problem.tex}
\input{04_goals.tex}
\input{05_structure.tex}
\input{06_related_work.tex}
\input{07_related_work_nat.tex}
\input{08_related_work_rtc.tex}
\input{09_related_work_signaling.tex}
\input{10_related_work_hypermedia.tex}
\input{11_related_work_collaboration.tex}
%\input{XX_implementation_model.tex}

@@
